---
title: "MTHM508_Coursework_II"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem

The homicides data set contains entries for all recorded homicides in London boroughs from 2003 to March 2022. The first of the month on which the homicide occurred is given as the recorded date. Year (counting the year in the data with 2003 = 1, 2004 = 2 etc), month (which month of the year) and season variables have been added to the data. The data also includes information on the age, sex and ethnicity of the victim, as well as whether the case was solved by the police or not. The method of killing is also recorded, as is whether or not the case involved domestic abuse.

```{r}
# Load libraries - suppress messages
suppressMessages(library(bayesplot))
suppressMessages(library(brms))
suppressMessages(library(reshape2))
suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(arm))
suppressMessages(library(boot))
# Read data
homicides = read.csv("homicides.csv")
# train test split
homicides_train = filter(homicides,year<15)
homicides_test = filter(homicides,year>=15)
#look at raw data
head(homicides_train, n=2)
```
Of the 20 years in the dataset, I will consider the first 15 as training and reserve the last 5 for testing.

```{r}
# train test split
homicides_train = filter(homicides,year<15)
homicides_test = filter(homicides,year>=15)
```
Some functions to 'binarize' the data are applied, as most of them are categorical (i.e. factors). If the crime was gun-related, it is grouped as "$1$" and if not, it is grouped as "$0$". This new variable is named "shooting".

```{r}
# Functions
#__________________________________________________________
# binarize the solved_status
binarize = function(df,col,set_to_one){
  df[col]=ifelse(df[col]==set_to_one,1,0)
  return(df)
}
# binarize the other categorical data
binarize_2 = function(df,col,set_to_one){
  df[col]=ifelse(df[col]==set_to_one,1,-1)
  return(df)
}
# convert ethnicity to num (modififed from code showcased during MTHM508 Pima Indians lectures)
ethinicity_num_conv = function(vec){
  ans <- as.character(vec)
  for(i in 1:length(vec)){
    if(vec[i] == 'Not Reported/Not known'){
      ans[i] = 1
    }
    else if(vec[i] == 'Other'){
      ans[i] = 2
    }
    else if(vec[i] == 'Black'){
      ans[i] = 3
    }
    else if(vec[i] == 'Asian'){
      ans[i] = 4
    }
    else{
      ans[i] = 5
    }
  }
  return(as.numeric(ans))
}
# Pipeline Function for cleaning and converting data types
pipeline = function(df){
  # convert month to num
  month_num = c()
  for(i in 1:nrow(df)){
    month_num = c(month_num,substr(df$recorded_date[i],1,1))
  }
  # Don't need recorded date -> year and month.name suffice
  df = tibble(subset(df, select = -c(recorded_date) ))
  df$eth_num = ethinicity_num_conv(df$observed_ethnicity)
  df$shooting = as.integer(df$method_of_killing=='Shooting')
  # binarize
  df = binarize(df=df, col='solved_status', set_to_one='Solved')
  df = binarize_2(df=df, col='sex', set_to_one='Male')
  df = binarize_2(df=df, col='domestic_abuse', set_to_one='Domestic Abuse')
  # convert age to num
  df$age_group_numeric = as.numeric(factor(df$age_group))
  return(df)
}

#__________________________________________________________
# transformations
homicides_1 = pipeline(homicides_train)
homicides_2 = pipeline(homicides_test)
# look at data
head(homicides_1,n=2)
```
This is a classfication task, so the model will be Logistic Regression. The model structure is

$$
y_{ij}|\eta_{ij},\theta \sim \mathcal{D}(g^{-1}(\eta_{ij}),\theta)
$$
where $g$ is the 'link function'. In this case, it is the $logit$ function, such that $g^{-1}(\eta)=\frac{1}{1+e^{-\eta}}$
The model for logistic regression dows not bear any $\theta$ parameter, so I will not use it going forward.


$$
\eta_j = b_0 + b_1 x_{1j}+b_2 x_{2j}+b_3x_{3j}+b_1x_{4j} + \beta_0 + \beta_1x_{1j}+\beta_2x_{2j}+\beta_3x_{3j}+\beta_1x_{4j}
$$

$$
\beta_j \sim \mathcal{N}(0,\Sigma)
$$
$$
\pi(b,\Sigma)
$$
Some factors that seem likely to influence the target variable can be sex, domestic abuse and whether the homicide was gun-related or not.

## Justification for Priors

**Intercept prior**

I look at the data to judge the level/ average rate of solving homicides, which is around 0.89. Opting for something weakly informative is advisable as there is no weight of literature behind setting these priors. As a result, choosing an intercept prior that is allowed to reach 0.89 (when put through the inverse logit function) as an extreme observation (~2 s.d.) can be reasonable. Thus, a normal distribution with mean $0$ can be allowed to have 2 s.d. = $logit(0.8944)=~2.137$. The following prior seems appropriate
$$\mathcal{N}(0,1.2)$$
This yields a 2 s.d of ~2.19.

The following are all originally categorical variables encoded as 1 and 0. Their priors will be similar to each other and different from the intercept prior.

**Sex prior**
As males are encoded as 1 and females are encoded as 0, a coefficient that is allowed to vary till 0.5 can be appropriate. Considering the intercept, this will roughly yield somewhere around $g^{-1}(2.19+0.5*1)=0.93$ and $g^{-1}(2.19-0.5*1)=0.84$, which is within bounds I would expect. Hence, following the same calculation as above, I choose a prior of $$\mathcal{N}(0,0.625)$$

**Domestic abuse prior**
With domestic abuse, I hypothesize that it is 'easier' to find the perpetrator. Living in the same household must greatly narrow down the search. That being said, I would not be comfortable with placing a very wide prior on the coefficient, as it can potentially take the response variable to above 0.99 chance of being solved - that seems unreasonable. Looking at the data, it appears that domestic abuse cases get solved ~96% of the time. $g(0.96)=3.18$. Considering the intercept of around 2.19, I can hit the maximum value of domestic abuse variable (i.e. 1) with ~0.98. Let's round that to 1 so that the distribution is allowed to go up to 1 roughly 95% of the time (i.e. ~2 s.d.)
This lands us with a prior 
$$\mathcal{N}(0,0.25)$$

**Shooting prior**
It can be reasoned that shootings make it relatively harder to solve the case as compared to physical assaults or stabbings. The latter two are more likely to leave behind DNA evidence or other identifiers at the scene, whereas a shooting can leave behind as little as a bullet shell (or casing), which may or may not be found at the crime scene. This is the reason why the shootings and non-shootings have been grouped separately, and are being considered in the first fit of the model.

As for the prior, I can believe the probability of solving slips down to ~75% if the crime was gun-related. In this scenario, $2.19-g^{-1}(0.75)=1.1$, i.e. the prior should be allowed to reach 1.3 in the 95th percentile (of its distribution). Therefore, $s.d. = \frac{1.1}{2}=0.55$. The prior is thus
$$\mathcal{N}(0,0.3025)$$
Priors for standard deviation of $\beta_j$ are next. Left alone, Stan would revert to a half-student_t(3, 0, 2.5). This can seem a little too wide. Suppose, for the intercept, which is already around 2.19, having a s.d. on top of that, reaching upto $\sqrt{2.5}*2=3.16$ would mean passing ~$2.19+3.16$ through the inverse logit function, bringing it very near to 1 (albeit this would be a highly improbable scenario). I would like to keep the s.d. priors tight around their original $b$ priors. A zero-mean normal distribution with small variance (~0.1) for the b priors and slightly larger (~0.5) for Intercept prior sounds plausible.

$$
\mathcal{N}(0,0.1) \ ; \ \mathcal{N}(0,0.5)
$$

```{r}
# Set priors
intercept_prior = set_prior('normal(0,1.2)',class='Intercept')
sex_prior = set_prior('normal(0,0.625)',class='b',coef='sex')
abuse_prior = set_prior('normal(0,0.4225)',class='b',coef='domestic_abuse')
shooting_prior = set_prior('normal(0,0.3025)',class='b',coef='shooting')

sd_priors_I <- set_prior("normal(0,0.5)",class="sd", group="observed_ethnicity", coef="Intercept")
sd_priors <- set_prior("normal(0,0.01)",class="sd", group="observed_ethnicity")

# Join priors into a vector
my_prior = c(intercept_prior,sex_prior,abuse_prior,shooting_prior,sd_priors_I, sd_priors)

fit_1 = brm(solved_status~sex+domestic_abuse+shooting
            +(shooting+domestic_abuse|observed_ethnicity),
          data=homicides_1,
          family=bernoulli,
          prior=my_prior,
          refresh=0)

summary(fit_1)
```
The eth_num variable does not seem to have a stable effect on the model. It will be discarded it in the next run. Trace plots are examined.
```{r}
plot(fit_1,ask=FALSE)
```
It appears that convergence has been reached.

```{r}
get_prior(solved_status~sex+domestic_abuse+shooting
            +(shooting+domestic_abuse|observed_ethnicity),
          data=homicides_1,
          family=bernoulli,
          prior=my_prior,)
```


